RELIKE: Reionization Effective Likelihood from Planck 2018 Data

### 2021-04-16 paper/rei.tex
Updates from Overleaf

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}

Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data. In light of the $\texttt{SRoll2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood. Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.

The code is publicly available on GitHub at \url{https://github.com/chenheinrich/RELIKE}. We publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model. If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space; we provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in \texttt{CosmoMC-relike}. We expect the code release to dramatically reduce the time for testing individual models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity function, star formation history, 21cm or line-intensity mapping in general, etc.

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
%
% longer Gaussian chains only 
%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
%
% longer KDE and Gaussian chains
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L. contours).  The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.   Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters 
which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains (still running)}
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f. adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects. [Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data? More importantly with data in general that wants to vary As during sampling? Put prior on As consistent with Planck?].

>>>>

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, adopts better foreground modeling techniques and allowing for improved reionization constraints.
%
%The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.
%\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
%
\ch{Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data. In this paper, we obtain new reionization PC constraints with this latest likelihood $\texttt{SRoll2}$. Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood code useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}. 
}

\ch{We publish this reionization effective likelihood code (RELIKE) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}. There are two packages that come with the code release: 1) we publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model. If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space. 2) We also provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in the package \texttt{CosmoMC-relike}. We expect that this code release would allow much simpler and faster testing of specific models in the Planck data, and to provide consistent and fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity functions, star formation history, 21cm and line-intensity mapping in general, etc.
}

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
%
% longer Gaussian chains only 
%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
%
% longer KDE and Gaussian chains
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L. contours).  The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.   Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters 
which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
suggested change in model - or find another reason for plotting the currently chosen one.
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f. adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects. \ch{[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data? More importantly with data in general that wants to vary As during sampling? Put prior on As consistent with Planck?]}.
